---
title: "Exercise 5 of \"Big Data Analytics\" (Winter Semester 2020/21)"
output: html_notebook
---

# Task 3: Outlier Detection

As for clustering, many outlier detection algorithms also use distances.
Thus, we should normalize the data:

```{r}
dataset <- data.frame(scale(faithful))
```

As a side note: Though the `faithful` dataset is rather simple, it even was used in the summary chapter of [a dissertation on outlier detection at our chair](http://dx.doi.org/10.5445/IR/1000120534).

## a) Visualization

You might remember the scatter plot from the clustering task:

```{r}
library(ggplot2)
ggplot(data = dataset) +
  geom_point(mapping = aes(x = eruptions, y = waiting)) +
  coord_fixed() # uniform axis scaling
```

We can plot univariate densities:

```{r}
ggplot(data = dataset) +
  geom_histogram(mapping = aes(x = eruptions))
ggplot(data = dataset) +
  geom_histogram(mapping = aes(x = waiting))
```

... but also bivariate densities:

```{r}
ggplot(data = dataset) +
  geom_bin2d(mapping = aes(x = eruptions, y = waiting), bins = 10) +
  scale_fill_distiller(palette = "Spectral")
```

With `plotly`, we can even make the plot interactive:

```{r}
library(plotly)
subplot(
  plot_ly(dataset, x = ~eruptions, color = I("black"), type = "histogram",
          name = "eruptions"),
  plotly_empty(type = "scatter", mode = "markers"), # arguments here serve no purpose other than suppressing warnings
  plot_ly(dataset, x = ~eruptions, y = ~waiting, type = "histogram2dcontour",
          showscale = FALSE, name = "faithful"),
  plot_ly(dataset, y = ~waiting, color = I("black"), type = "histogram",
          name = "waiting"),
  nrows = 2, heights = c(0.2, 0.8), widths = c(0.8, 0.2),
  shareX = TRUE, shareY = TRUE
)
```

Okay, enough plotting for now.
In our case, the density plots are not really helpful in finding prominent outliers.
We can clearly see the two clusters, even in univariate plots, and decreasing density far from the centers.
However, this is what we expect from roughly normally distributed data.
There are no points which are really far away from the rest.
In the scatter plots, we see that there are some points with a rather high distance to the two cluster centers and a rather high distance to other points.
"Rather" is a rather fuzzy term, though.
Different outlier detection methods try to capture notions of outlierness formally in different ways.
However, there often remains the challenge of finding a suitable threshold:
When is a point an inlier and when is it an outlier?
The definition might depend on the use case rather than some fixed statistical boundaries.
As a remedy, many outlier detection algorithms do not return a binary decision, but a continuous outlier score.
This allows moving the threshold flexibly.

## b) Outlier Detection

Both outlier detection function proposed in the task return a vector of outlier scores.
We define a function for creating a scatter plot with outliers highlighted:

```{r}
plotOutliers <- function(outlierScores) {
  plotData <- cbind(dataset, OutlierScore = outlierScores)
  ggplot(data = plotData) +
    geom_point(mapping = aes(x = eruptions, y = waiting, color = OutlierScore,
                             size = OutlierScore)) +
    coord_fixed() + # uniform axis scaling
    scale_color_distiller(palette = "Spectral") +
    guides(color = guide_legend(), size = guide_legend()) # joint legend
}
```

From the implementation perspective, we basically add the outlier scores as another column to the dataset.
Next, we use the aesthetics `color` and `size` to style points based on their outlier score.
Let's see the function in action:

```{r, fig.height = 4, fig.width = 8}
gridExtra::grid.arrange(
  plotOutliers(DDoutlier::KNN_SUM(dataset, k = 5)) + ggtitle("kNN"),
  plotOutliers(dbscan::lof(dataset, k = 5)) + ggtitle("LOF"),
  nrow = 1, respect = TRUE
)
```

Another implementation detail:
Our plot function returns the plots as objects, which we arrange side-by-side.

We can see some commonalities and some differences between the results of the two methods.
For example, points in the centers of the clusters generally have low outlier scores.
The top-right-most point has a high outlier score for both methods.

In contrast, several points in the region between the clusters have a high score for kNN, but a low score for LOF.
These points are in a region of low density, with comparatively high distances to nearest neighbors.
For kNN, only the distance matters, and the same distance scale is considered globally.
For LOF, points which have a high kNN distance, but also neighbors with a high kNN distance, get a comparatively low score.
This means the local situation plays a bigger role.

In the bottom-left corner, you can see two points with high LOF score, but medium kNN score.
For kNN, these points clearly have a higher kNN distance than points in the cluster center.
However, they are not far off, resulting in a comparatively low outlier score.
For LOF, the neighbors of these points have a clearly lower distance to *their* neighbors, because they are closer to the center.
Thus, the LOF score of the two points under discussion is rather high.

## c) Subspace Outliers

To ease comparison, we create a grid with univariate outlier scores, bivariate outlier scores and bivariate density.

```{r, fig.height = 8, fig.width = 8}
gridExtra::grid.arrange(
  plotOutliers(DDoutlier::KNN_SUM(dataset$eruptions, k = 5)) +
    ggtitle("Outliers for \"eruptions\"") + theme(legend.position = "none"),
  plotOutliers(DDoutlier::KNN_SUM(dataset$waiting, k = 5)) +
    ggtitle("Outliers for \"waiting\"") + theme(legend.position = "none"),
  plotOutliers(DDoutlier::KNN_SUM(dataset, k = 5)) +
    ggtitle("Outliers for full space") + theme(legend.position = "none"),
  ggplot(data = dataset) +
    geom_bin2d(mapping = aes(x = eruptions, y = waiting), bins = 10) +
    scale_fill_distiller(palette = "Spectral") +
    ggtitle("Density") + theme(legend.position = "none"),
  nrow = 2, ncol = 2, respect = TRUE
)
```

There definitely are points whose outlier score differs between subspaces.
For example, the points with the lowest and highest waiting times have a high outlier score regarding `waiting`,
but a lower score regarding `eruptions`.
Also, there is a point in the middle at roughly (-0.4, -0.1),
which is quite outlying regarding `eruptions` (it is the only point in some "vertical corridor" around it),
but rather normal regarding `waiting` (some other points are in same "horizontal corridor" around it).
There are also non-trivial outliers.
For example, the point at roughly (-1, 0) is perfectly normal in either one-dimensional space,
but has a high outlier score in the two-dimensional space.
